---
title: rice
output:
  html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{rice}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# Introduction

Radiocarbon dating requires a range of calculations, for example calibration[^1][^2][^3][^4], translations between pMC, F14C, C14 age and D14C, and assessing the impacts of contamination. This package provides functions to do so in R. 

# Installation

On first usage of the package, it has to be installed: 

```{r, eval=FALSE}
install.packages('rice')
``` 

The companion data package 'rintcal' which has the radiocarbon calibration curves will be installed if it isn't already. New versions of R packages appear regularly, so please re-issue the above command regularly to remain up-to-date, or use:

```{r, eval=FALSE}
update.packages()
``` 

To obtain access to the calibration curves and radiocarbon functions, first the package has to be loaded:

```{r}
library(rice)
```

# Calibration curves

Calibration curves can be plotted:

```{r, fig.width=4, fig.asp=.8}
draw.ccurve()
```

Or, comparing two calibration curves:

```{r, fig.width=4, fig.asp=.8}
draw.ccurve(1000, 2020, BCAD=TRUE, cc2='marine20', add.yaxis=TRUE)
```

Or zooming in to between AD 1600 and 2000 (using the BCAD scale):

```{r, fig.width=4, fig.asp=.8}
draw.ccurve(1600, 1950, BCAD=TRUE)
```

Interesting things happened after 1950, as can be seen by adding a postbomb curve:

```{r, fig.width=4, fig.asp=.8}
draw.ccurve(1600, 2020, BCAD=TRUE, cc2='nh1')
```

The postbomb curve dwarfs the IntCal20 curve, so we could also plot both on separate vertical axes:

```{r, fig.width=4, fig.asp=.8}
draw.ccurve(1600, 2020, BCAD=TRUE, cc2='nh1', add.yaxis=TRUE)
```

# Realms

This package also provides functions to translate values between the radiocarbon-relevant 'realms' of cal BP (calendar years before AD 1950), BC/AD, C14, F14C, pMC and D14C. The following Table lists the available functions:

| from\\to  | calBP         | BCAD          | C14          | F14C         | pMC         | D14C         |
|-----------|---------------|---------------|--------------|--------------|-------------|--------------|
| **calBP** | -             | `calBPtoBCAD` | `calBPtoC14` | `calBPtoF14C`| `calBPtopMC`| `calBPtoD14C`|
| **BCAD**  | `BCADtocalBP` | -             | `BCADtoC14`  | `BCADtoF14C` | `BCADtopMC` | `BCADtoD14C` |
| **C14**   | NA            | NA            | -            | `C14toF14C`  | `C14topMC`  | `C14toD14C`  |
| **F14C**  | NA            | NA            | `F14CtoC14`  | -            | `F14CtopMC` | `F14CtoD14C` |
| **pMC**   | NA            | NA            | `pMCtoC14`   | `pMCtoF14C`  |  -          | `pMCtoD14C`  |
| **D14C**  | NA            | NA            | `D14CtoC14`  | `D14CtoF14C` | `D14CtopMC` |  -           |


As an example of the above functions, the IntCal20 C14 age and error belonging to one or more cal BP or BCAD ages can be found (interpolating linearly where necessary):  

```{r}
calBPtoC14(10.5)
BCADtoC14(1940:1950)
```

To translate between BC/AD and cal BP ages, we can use (the last example avoids 0 BC/AD):

```{r}
BCADtocalBP(2024)
BCADtocalBP(-1, zero=TRUE)
BCADtocalBP(-1, zero=FALSE)
```

D14C ($\Delta$^14^C, a proxy for atmospheric ^14^C concentration at *t* cal BP) can be transferred to F^14^C, and the other way around:

```{r}
D14CtoF14C(152, t=4000)
F14CtoD14C(0.71, t=4000)
```

This can also be done with C14 ages:

```{r}
C14toD14C(0.71, t=4000)
D14CtoC14(152, t=4000)
```

These functions can be used to investigate $\Delta^{14}C$ over time:

```{r, fig.width=4, fig.asp=.8}
x <- seq(0, 55e3, length=1e3)
cc <- calBPtoC14(x)
Dcc <- calBPtoD14C(x)

par(mar=c(4,3,1,3), bty="l")
plot(x/1e3, Dcc[,1]+Dcc[,2], type="l", xlab="kcal BP", ylab="")
mtext(expression(paste(Delta, ""^{14}, "C")), 2, 1.7)
lines(x/1e3, Dcc[,1]-Dcc[,2])

par(new=TRUE)
plot(x/1e3, (cc[,1]-cc[,2])/1e3, type="l", xaxt="n", yaxt="n", col=4, xlab="", ylab="")
lines(x/1e3, (cc[,1]+cc[,2])/1e3, col=4)
mtext(expression(paste(""^{14}, "C kBP")), 4, 2, col=4)
axis(4, col=4, col.axis=4, col.ticks=4)
```


# Contamination

To calculate the effect of contamination on radiocarbon ages, e.g. what age would be observed if material with a "true" radiocarbon age of 5000 +- 20 ^14^C BP would be contaminated with 1% of modern carbon (F^14^C=1)?

```{r}
contaminate(5000, 20, .01, 1)
```

Or imagine that you dated a dino bone, which is much much beyond the limit of radiocarbon dating:

```{r}
contaminate(66e6, 1e6, .01, 1, F14C.er=.01)
```

The effect of different levels of contamination can be visualised:

```{r, fig.width=6, fig.asp=.8}
real.14C <- seq(0, 50e3, length=200)
contam <- seq(0, .1, length=101) # 0 to 10% contamination
contam.col <- rainbow(length(contam))
plot(0, type="n", xlim=c(0, 55e3), xlab="real 14C age", ylim=range(real.14C), ylab="observed 14C age")
for(i in 1:length(contam))
  lines(real.14C, contaminate(real.14C, c(), contam[i], 1, decimals=5), col=contam.col[i])
contam.legend <- seq(0, .1, length=6)
contam.col <- rainbow(length(contam.legend)-1)
text(50e3, contaminate(50e3, c(), contam.legend, 1), 
  labels=contam.legend, col=contam.col, cex=.7, offset=0, adj=c(0,.8))
```

If that is too much code for you, try this function instead:

```{r, fig.width=6, fig.asp=.8}
draw.contamination()
```

# Fractions

Sometimes, one needs to estimate a missing radiocarbon age from a sample which has C14 dates on both the entire sample and on fractions, but where one of the samples was too small to be dated. This can be used in for example soils separated into size fractions, where one of the samples turns out to be too small to be dated. Requires to have the bulk age, the ages of the dated fractions, and the carbon contents and weights of all fractions.

```{r}
Cs <- c(.02, .05, .03, .04) # carbon contents of each fraction
wghts <- c(5, 4, 2, .5) # weights for all fractions, e.g., in mg
ages <- c(130, 130, 130, NA) # ages of all fractions. The unmeasured one is NA
errors <- c(10, 12, 10, NA) # errors, unmeasured is NA
fractions(150, 20, Cs, wghts, ages, errors) # assuming a bulk age of 150 +- 20 C14 BP 
```

# Calibration

Now on to calibration of radiocarbon dates. We can obtain the calibrated probability distributions from radiocarbon dates, e.g., one of 130 &pm; 10 C14 BP:

```{r, fig.width=4, fig.asp=.8}
calib.130 <- caldist(130, 10, BCAD=TRUE)
plot(calib.130, type="l")
```

It is also possible to find the likelihood of a single calendar year for our radiocarbon age, e.g., 145 cal BP:

```{r}
l.calib(145, 130, 10)
```

For reporting purposes, calibrated dates are often reduced to their 95% highest posterior density (hpd) ranges (please report all, not just your favourite one!):

```{r}
hpd(calib.130)
```

Additionally, calibrated dates are often reduced to single point estimates. Note however how poor representations they are of the entire calibrated distribution!

```{r, fig.width=4, fig.asp=.8}
calib.2450 <- caldist(2450, 20)
plot(calib.2450, type="l")
points.2450 <- point.estimates(calib.2450)
points.2450
abline(v=points.2450, col=1:4, lty=2)
```

Want a plot of the radiocarbon and calibrated distributions, together with their hpd ranges?
```{r, fig.width=5, fig.asp=1}
calibrate(2450, 20)
```

Sometimes one would want to smooth a calibration curve to take into account the fact that material has accumulated over a certain time (e.g., a tree, or peat). To do so, a calibration curve can be smoothed to produce a tailor-made calibration curve, after which this one is used to calibrate the date:

```{r, fig.width=5, fig.asp=1}
mycurve <- smooth.ccurve(smooth=50)
calibrate(2450, 20, thiscurve=mycurve)
```

Calibrating 'young' radiocarbon dates (close to 0 C14 BP) can cause an error, because a bomb curve might be required to capture the youngest ages. Do not worry, there is an option to avoid that error:

```{r, fig.width=5, fig.asp=1}
try(calibrate(130,30))
calibrate(130, 30, bombalert=FALSE)
```

It is also possible to analyse the calibrated probability distributions, e.g. what is the probability (between 0 and 1) that the date stems from material that is of the age of 150 cal BP or younger? Or that it is older than that age?

```{r}
younger(150, 130, 10)
older(150, 130, 10)
```


# Marine offsets

Dates on marine material will often have to be calibrated with the Marine20 calibration curve[^5], and many coastal locations will have an additional regional reservoir offset (deltaR) (Reimer and Reimer 2006)[^6]. The on-line database at <http://calib.org/marine/> is very useful for this; it features the radiocarbon ages and deltaR of many shells of known collection date. The data from this database were downloaded (in August 2024) and can be queried. For example, a map can be drawn with all shell data within certain coordinates:

```{r, fig.width=5, fig.asp=1}
myshells <- map.shells(S=54, W=-8, N=61, E=0) # the northern part of the UK
```

The output can also be queried:

```{r, fig.width=5, fig.asp=1}
head(myshells)
shells.mean(myshells)
```

You can also extract say the 20 shells closest to a coordinate, e.g., 120 East and 10 North:
```{r, fig.width=5, fig.asp=1}
myshells <- find.shells(120, 10, 20)
shells.mean(myshells, distance=TRUE)
```

# Multiple calibrations

You can also draw one or more calibrated distributions:

```{r, fig.width=4, fig.asp=1}
set.seed(123)
dates <- sort(sample(500:2500,5))
errors <- .05*dates
depths <- 1:length(dates)
my.labels <- c("my", "very", "own", "simulated", "dates")
draw.dates(dates, errors, depths, BCAD=TRUE, labels=my.labels, age.lim=c(0, 1800))
```

or add them to an existing plot:

```{r, fig.width=4, fig.asp=1}
plot(300*1:5, 5:1, xlim=c(0, 1800), ylim=c(5,0), xlab="AD", ylab="dates")
draw.dates(dates, errors, depths, BCAD=TRUE, add=TRUE, labels=my.labels, mirror=FALSE)
```

or get creative (inspired by Jocelyn Bell Burnell[^7], Joy Division[^8] and the Hallstatt Plateau[^9]):

```{r, fig.width=4, fig.asp=1}
par(bg="black", mar=rep(1, 4))
n <- 50; set.seed(1)
draw.dates(rnorm(n, 2450, 30), rep(25, n), n:1,
  mirror=FALSE, draw.base=FALSE, draw.hpd=FALSE, col="white",
  threshold=1e-28, age.lim=c(2250, 2800), ex=.8)
```

[^1]: Stuiver, R., Polach, H.A., 1977. Discussion: reporting of 14C data.
*Radiocarbon* 19, 355-363 <http://dx.doi.org/10.1017/S0033822200003672>

[^2]: Reimer, P.J., Brown, T.A., Reimer, R.W., 2004. Discussion: reporting
and calibration of post-bomb 14C Data. *Radiocarbon* 46, 1299-1304 <http://dx.doi.org/10.1017/S0033822200033154>

[^3]: Millard, R., 2014. Conventions for reporting radiocarbon determinations. *Radiocarbon* 56, 555-559 <http://dx.doi.org/10.2458/56.17455>

[^4]: Reimer, P.J., et al., 2020. The IntCal20 Northern Hemisphere radiocarbon age calibration curve (0-55 cal kBP). *Radiocarbon* 62, 725-757 <http://dx.doi.org/10.1017/S0033822200032999>

[^5]: Heaton, T.J., et al., 2020. Marine20-the marine radiocarbon age calibration curve (0-55,000 cal BP). *Radiocarbon* 62, 779-820 <http://dx.doi.org/10.1017/RDC.2020.68>

[^6]: Reimer, P.J., Reimer, R.W., 2006. A marine reservoir correction database and on-line interface. *Radiocarbon* 43, 461-463. <http://dx.doi.org/10.1017/S0033822200038339>

[^7]: https://www.cam.ac.uk/stories/journeysofdiscovery-pulsars

[^8]: https://www.radiox.co.uk/artists/joy-division/cover-joy-division-unknown-pleasures-meaning/

[^9]: https://en.wikipedia.org/wiki/Hallstatt_plateau
